---
title: "Week 3 - Homework"
format: 
  html: default
date: "2025-11-19"
date-format: " "
categories: 
  - week03
  - homework
editor_options: 
  chunk_output_type: console
---

1. Create a matrix of two columns to store numbers 1 to 8,
use the `dimnames` parameter of the `matrix` function to name
the rows and columns directly. (see `?matrix`)
(You can name the rows/columns whatever you want.)

```{r}

```


Use the following code to import the file "read-counts.csv" 
(you have already downloaded it for the hand-on examples of week01).
Name the imported data `expr_data`.

```{r}
expr_data <- read.table(
  file = "../exos_data/read-counts.csv",  # replace the path with your own
  header = TRUE, sep = ",", row.names = 1
)

```

2. Create a list `info_expr` to store information about `expr_data`:  
  - "col_names": column names
  - "row_names": row names
  - "dimension": number of rows and columns
  - "sum_sample": the sum of expression level across all genes for each sample
  - "avg_gene": the average expression level across all samples for each gene

```{r}

```

3. Calculate the median of the average gene expression levels across all samples.

```{r}

```

4. Use `summary()` to generate summary statistics of the sum of expression level across all genes.

```{r}

```

5. Calculate the log2 fold change of the gene *TOS6* for the WT and RRP6 samples.
Is *TOS6* up or down-regulated in the RRP6 samples compared to WT?

```{r}

```

6. Perform a Wilcoxon test to check if the *TOS6* expression is different between
WT and RRP6 samples. 

  - Store the test results in a variable (object) called `res_wilcox`.
  What is the data structure of the results? Extract the p-value from the results.
  - Use $\alpha = 0.05$, what is your conclusion?

```{r}

```


7. Visualize the expression level of *TOS6* for WT and RRP6 samples using boxplot.

```{r}

```


The homework correction is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week03/.r_w03_work_corr.R){target='_blank'}