---
title: "Week 1 - Homework"
format: 
  html: default
categories: 
  - week01
  - homework
editor_options: 
  chunk_output_type: console
---

Note: There may be other valid ways to solve this exercise — this is just one possible solution!

1. Import the data file "**visceral_fat.csv**".

  - First, download the `visceral_fat.csv` file via this [link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/exos_data/visceral_fat.csv).
  - Then, upload the file to a folder of your choice within your RStudio project.
  - Next, import the data into RStudio and name the data frame `visceral_fat`. 

Tip:

Use the **click-button method** in RStudio to upload and import the file. Then, copy and paste the R commands generated into a code chunk in your Quarto report.

Once imported, you can use the `visceral_fat` dataset for further analysis in your report.

```{r}
# command from the import pop-up window
visceral_fat <- read.csv("../exos_data/visceral_fat.csv") # replace the path by your own
```

2. Summarize the dataset.

Run the function `summary(visceral_fat)` to generate a summary of the dataset. Describe what information you obtained.

```{r}
summary(visceral_fat)
```

This function returns the minimum, maximum, mean, and quartile values for each column.
The column `ethnicity` contains one missing value (`NA`).

3. Draw a histogram.

Create a histogram for the diastolic blood pressure using the column `visceral_fat[["diastolic_bp__mmhg"]]`. Describe what you observe from the plot.

```{r}
hist(visceral_fat[["diastolic_bp__mmhg"]])
```

The histogram shows a skewed distribution with some extreme values. Most of the values are concentrated in the bins on the left, while a single bin far on the right suggests outliers.

4. Handle an outlier.

After consulting with a clinician, you discover that the 27th value of the diastolic blood pressure was incorrectly recorded.

  - Ignore this value using `visceral_fat[["diastolic_bp__mmhg"]][-27]`
  - Redraw the histogram after excluding the outlier and describe any changes you observe.

```{r}
hist(visceral_fat[["diastolic_bp__mmhg"]][-27])
```

The histogram becomes more symmetrical after removing the outlier.
The most frequent values fall within the range of 65 to 70.

5. Compare mean and median.

Calculate the mean (`mean()`) and the median (`median()`) for the diastolic blood pressure

  - before ignoring the 27th value: `visceral_fat[["diastolic_bp__mmhg"]]` 
  - after ignoring the 27th value: `visceral_fat[["diastolic_bp__mmhg"]][-27]`
  
Describe how the outlier affects the mean and median.

```{r}
mean(visceral_fat[["diastolic_bp__mmhg"]])
median(visceral_fat[["diastolic_bp__mmhg"]])
mean(visceral_fat[["diastolic_bp__mmhg"]][-27])
median(visceral_fat[["diastolic_bp__mmhg"]][-27])
```

The median remains unchanged.

The mean decreases after removing the outlier,
highlighting its sensitivity to extreme values.
The median, in contrast, is robust to extreme values.

6. Click "Render" to generate your Quarto report.
