---
title: "Week 1 - Homework"
format: 
  html: default
date: "2025-11-05"
date-format: " "
categories: 
  - week01
  - homework
editor_options: 
  chunk_output_type: console
---

Write your code in a new R script.
You can use `#` at the beginning of a line to add a comment in your script.

1. Import the data file "**visceral_fat.csv**".

  - First, download the `visceral_fat.csv` file via this [link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/exos_data/visceral_fat.csv).
  - Then, upload the file to a folder of your choice within your RStudio project.
  - Next, import the data into RStudio and name the data frame `visceral_fat`. 

Tip:

Use the **click-button method** in RStudio to upload and import the file. Then, copy and paste the R commands generated into your script.

Once imported, you can use the `visceral_fat` dataset for further analysis in your report.

```{r}

```

2. Summarize the dataset.

Run the function `summary(visceral_fat)` to generate a summary of the dataset. Describe what information you obtained.

```{r}

```

3. Draw a histogram.

Create a histogram for the diastolic blood pressure using the column `visceral_fat[["diastolic_bp__mmhg"]]`. Describe what you observe from the plot.

```{r}

```

4. Handle an outlier.

After consulting with a clinician, you discover that the 27th value of the diastolic blood pressure was incorrectly recorded.

  - Ignore this value using `visceral_fat[["diastolic_bp__mmhg"]][-27]`
  - Redraw the histogram after excluding the outlier and describe any changes you observe.

```{r}

```

5. Compare mean and median.

Calculate the mean (`mean()`) and the median (`median()`) for the diastolic blood pressure

  - before ignoring the 27th value: `visceral_fat[["diastolic_bp__mmhg"]]` 
  - after ignoring the 27th value: `visceral_fat[["diastolic_bp__mmhg"]][-27]`
  
Describe how the outlier affects the mean and median.

```{r}


```


The homework correction is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week01/.r_w01_work_corr.R)