---
title: "Week 5 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2025-12-02"
date-format: " "
categories: 
  - week05
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week05/r_w05_exos.R){target='_blank'}

## Goals

- Install new R packages
- Create basic plots with ggplot2

## Install New Packages

We will use a couple of additional R packages for this training
in session 9, for example:

- {[tidyr](https://tidyr.tidyverse.org){target='_blank'}}: provides functions that help you get to tidy data
- {[dplyr](https://dplyr.tidyverse.org){target='_blank'}}: provides functions that help you to manipulate data 


1. Install the {`dplyr`} and {`tidyr`} packages.

```{r}
# install.packages("dplyr") # you only need to install it once
# install.packages("tidyr") # you only need to install it once
```

2. Check if you can find `{dplyr}` and `{tidyr}` in the "*Packages*" panel in RStudio.

3. Please use the following code to create a data frame of all installed packages in your system.
Show a couple of first lines of the data frame.

```{r}
my_pkgs <- as.data.frame(installed.packages()[, c(1, 3:4)])
head(my_pkgs)
```

4. According to `my_pkgs`, how many packages are installed in your system?

```{r}
nrow(my_pkgs) # the number can be different
```

5. Extract the rows for the "ggplot2", "tidyr", "dplyr" packages from `my_pkgs`.

```{r}
# extract by rownames
my_pkgs[c("ggplot2", "tidyr", "dplyr"), ]

## or
# extract by filtering the names in the "Package" column
my_pkgs[my_pkgs$Package %in% c("ggplot2", "tidyr", "dplyr"), ]
```


## Create a Basic Histogram

In the hands-on examples of session 1 and 2, we have drawn histograms with the basic R function `hist()`.

Now let's try with functions from {`ggplot2`} package.

0. Import the `read-counts.csv` file into RStudio and name the data as `counts`.

```{r}
library(readr)
counts <- read_csv("../exos_data/read-counts.csv") # replace the path by the yours
dim(counts)
```


1. Load the {`ggplot2`} package.
Then create a histogram with the function `geom_histogram()` 
for all genes of the sample "WT.2".
Have you noticed the message from `geom_histogram()` about the bins? 

```{r}
library(ggplot2)
ggplot(data = counts, aes(x = WT.2)) +
  geom_histogram()
```

2. Create a new histogram for the sample "WT.2",
but this time use the log2 of the gene counts.
Do you see the messages from `geom_histogram()`? 

```{r}
ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram()
```

3. Try 10 and 20 for the `bins` parameter of `geom_histogram()`.
Observe how the histogram changes.

```{r}
ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram(bins = 10) +
  labs(title = "10 bins")

ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram(bins = 20) +
  labs(title = "20 bins")
```

4. We will use 10 bins for the histogram.
Change the color of the bars to blue.

```{r}
ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram(bins = 10, color = "blue")
```

5. Based on the previous figure, fill the bars in orange.

```{r}
ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram(bins = 10, color = "blue", fill = "orange")
```

6. Based on the previous figure, change:

- x-axis title to be: "log2 gene expression"
- y-axis title to be: "Counts"
- plot title to show sample name and the number of genes used.

```{r}
ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram(bins = 10, color = "blue", fill = "orange") +
  labs(
    x = "log2 gene expression",
    y = "Counts",
    title = paste0(nrow(counts), " genes of WT.2 sample")
  )
```

7. Based on the previous figure, use the theme `theme_bw()`.

```{r}
ggplot(data = counts, aes(x = log2(WT.2))) +
  geom_histogram(bins = 10, color = "blue", fill = "orange") +
  labs(
    x = "log2 gene expression",
    y = "Counts",
    title = paste0(nrow(counts), " genes of WT.2 sample")
  ) +
  theme_bw()
```


## Create a Boxplot with ggplot

In the hands-on examples of session 3,
we demonstrated that there is a significant difference in
the expression of *LOH1* between WT and SET1 samples.

This gene is upregulated in SET1 samples compared to WT samples,
the log2 fold change is 2.85.

Now, let's draw a more sophisticated boxplot with ggplot2.

0. Prepare a data frame for the boxplot. We need:

- a column for the expression levels of different samples
- a column to indicate what is the sample group (WT or SET1)

```{r}
gg_df <- data.frame(
  expr_value = c(
    unlist(counts[counts$Feature == "LOH1", paste0("WT.", 1:10)]),
    unlist(counts[counts$Feature == "LOH1", paste0("SET1.", 1:10)])
  ),
  gp = rep(c("WT", "SET1"), each = 10)
)
gg_df
```

1. Create a boxplot for the expression level by sample group.

```{r}
p_base <- ggplot(gg_df, aes(x = gp, y = expr_value)) +
  geom_boxplot()

p_base
```

2. Modify labels:

- Remove x-axis title
- Change y-axis title to "Expression Level"
- Add a plot title "Expression of LOH1 in SET1 and WT samples"
- Add a subtitle "log2FoldChange = 2.85"

```{r}
p_base <- p_base + labs(
  x = NULL,
  y = "Expression Level",
  title = "Expression of LOH1 in SET1 and WT samples",
  subtitle = "log2FoldChange = 2.85"
)
p_base
```

3. Add a layer of scatter plot over the boxplot.

- Color the points by sample group.
- Use the `alpha` parameter to let points be semi-transparent.

```{r}
p_base <- p_base +
  geom_point(
    aes(color = gp),
    alpha = 0.5
  )
p_base
```

4. Change the theme to `theme_minimal()`. 

```{r}
p_base <- p_base + theme_minimal()
p_base
```

5. With the `theme()` function:

- Move the plot title to the center.
- Hide the legend.

```{r}
p_base <- p_base + theme(
  plot.title = element_text(hjust = 0.5), # center plot title
  legend.position = "none"
)
p_base
```


---

#### Good job! ðŸ‘ðŸ‘ You've taken your first big steps into ggplot2, keep it up! 


