---
title: "Week 2 - Homework"
format: 
  html: default
date: "2025-11-12"
date-format: " "
categories: 
  - week02
  - homework
editor_options: 
  chunk_output_type: console
---

Use the following code to import the file "read-counts.csv" 
(you have already downloaded it for the hand-on examples of week01).
Name the imported data `expr_data`.

```{r}
expr_data <- read.table(
  file = "../exos_data/read-counts.csv",  # replace the path with your own
  header = TRUE, sep = ",", row.names = 1
)
```

1. Is the imported data a matrix?

```{r}

```

2. Transform it to a matrix using `as.matrix()`.

```{r}

```

3. What are the dimensions of the matrix?

```{r}

```

4. What are the column names and row names?

```{r}

```


5. Extract all genes' expression for the 10 WT samples (WT.1, WT.2, ..., WT.10),
and store it in a variable called `expr_wt`.
Show the first few rows using `head()` function.

```{r}

```

6. What is the average expression level across all WT samples for each gene?
Store the results in a variable called `avg_wt`.

```{r}

```

7. Draw a histogram to check the distribution of the average gene expression level for WT samples.

```{r}

```

8. Repeat questions 5 to 7 for the SET1 samples (SET1.1, SET1.2, ..., SET1.10).

```{r}

```

9. What is the absolute difference between the average expression of WT and SET1 samples for each gene? 
Which gene shows the biggest difference?

::: {.callout-tip collapse="true" title="Hints"}
- Use `abs()` to get the absolute value. (`?abs`)
- The `max()` function helps to find the maximum value from a vector. (`?max`)
- The `which.max()` function helps to find the index of the maximum value from a vector. (`?which.max`)
:::

```{r}

```


10. Compare the following matrix to the one we created during the course' exercise (the matrix named `expression_levels`).
What is the key difference in how the matrix is constructed? 
Provide a brief explanation.

```{r}
matrix(
  c(10, 12, 15, 20, 18, 22, 14, 16, 19, 8, 9, 7, 25, 30, 28),
  ncol = 3
)
```


The homework correction is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week02/.r_w02_work_corr.R)
