---
title: "Week 7 - Homework"
format: 
  html: default
date: "2025-03-12"
date-format: " "
categories: 
  - week08
  - homework
editor_options: 
  chunk_output_type: console
---

Write your code directly in a Quarto document.

To create a Quarto document: go to **File** -> **New File** -> **Quarto Document**, then click **Create**.

## Mini Data Project (Part 2)

We already saw how to import and tidy the data for structure 1 during the exercise.

Just a reminder, the expression of 20 genes was measured in mice aged from 10 to 60 days, with both male and female mice, and each group consists of 6 animals. The gene expression data is stored in two files, one for each brain structure. Within each file, rows represent the different ages, and columns represent the gene, sex, and animal.

The data is available in two CSV files:

- [data_anonym_struc1_noise.csv](../exos_data/data_anonym_struc1_noise.csv)
- [data_anonym_struc2_noise.csv](../exos_data/data_anonym_struc2_noise.csv)

1. Use `lapply()` to import and reshape the two data files at once. Don't forget to change the necessary parameters.

Hints:

- Create a vector to store the data files' path.
- Use `lapply()` to repeat the import and reshape steps on the created path vector.
- Bind rows to have a final data frame for both structures. (Use `dplyr::bind_rows()` or `rbind()`)

```{r}
library(tidyverse)

# create a vector for paths
data_paths <- c(
  "../exos_data/data_anonym_struc1_noise.csv",
  "../exos_data/data_anonym_struc2_noise.csv"
)

# import all files
data_list <- lapply(seq_along(data_paths), function(i) {
  readr::read_csv2(
    file = data_paths[i],
    locale = locale(decimal_mark = ",")
  ) |>
  rename(age = ...1) |>
  pivot_longer(cols = -1, names_to = "id", values_to = "value") |>
  mutate(struc = paste0("s", i)) |>
  extract(
    col = id,
    into = c("gene_id", "sex", "animal"),
    regex = "([0-9]+)([MF])([A-F])"
  )
})

# bind data frames by rows
all_data <- bind_rows(data_list[[1]], data_list[[2]])
all_data
# or use the do.call() to automate the binding (?do.call)
all_data <- do.call(
  what = bind_rows, # the function name
  args = data_list # the list
)

```

2. Remove lines with NA in expression value.

```{r}
all_data <- all_data |> drop_na()
```

3. At age 60 and in the brain structure 2, which animal has the smallest expression level in each sex for gene 2 and 16?

```{r}
all_data |>
  filter(age == 60 & struc == "s2" & gene_id %in% c("2", "16")) |>
  group_by(sex, gene_id) |>
  arrange(value) |>
  slice(1)
```

4. Explore how to use `facet_grid()` in ggplot2 to display figures in separate facets.
Create a boxplot to visualize the expression levels of gene 16 for both sexes at different age time points, with each brain structure shown in its own facet.

```{r}
gene_i <- "16"
all_data |>
  filter(gene_id == gene_i) |>
  ggplot(aes(x = as.character(age), y = value)) +
  geom_boxplot(aes(color = sex)) +
  labs(
    x = "Age (days)", y = "Expression Level", color = "Sex",
    title = paste("Gene", gene_i)
  ) +
  facet_grid(rows = vars(struc)) +
  theme_light()
```


---

Click "Render" to generate your Quarto report.
