---
title: "Week 7 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2026-01-13"
date-format: " "
categories: 
  - week07
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week07/r_w07_exos.R)

## Goals

- Clean, reshape, and manipulate real-world data using {`dplyr`} and {`tidyr`} 
- Integrate tidy data into downstream analysis (e.g., for visualization or statistical analysis)

## Install {tidyverse} and Load the Package

```{r}
# install.packages("tidyverse")
library(tidyverse)
```


## Mini Data Project

This mini data project is based on a real project that focuses on gene expression across different time points.

A researcher has measured the expression levels of **20 genes** (anonymed as 1 to 20) using the RT-qPCR technique.
The gene expression was assessed in **two structures** of the mouse brain.
Mice ranged in age from 10 to 60 days (**10, 15, 20, 25, 30, 35, 40, 45, 50, 60 days**),
and the experiment was repeated with both **male** and **female** mice,
with **6 animals** (named from A to F) in each group.

According to the researcher, the data was stored in two files, one for each brain structure.
Within each file, **rows represent the different ages,
and columns represent the gene, sex, and animal.**
For example, column `1MA` represents the expression level of *gene 1* from the *male* mouse *A*,
column `15FD` represents the expression level of *gene 15* from the *female* mouse *D*.

A small Gaussian noise has been added to the original data, preserving the overall structure.

The data is available in two CSV files:

- [data_anonym_struc1_noise.csv](../exos_data/data_anonym_struc1_noise.csv)
- [data_anonym_struc2_noise.csv](../exos_data/data_anonym_struc2_noise.csv)

We will focus on the data from the brain structure 1.
**The aim is to perform an exploratory analysis to have a glimpse of the relationship between variables.**

### Import the Data

1. Please download the `data_anonym_struc1_noise.csv` file.
Observe your data file:

- Is there a header line? 
- What is the separator between columns?
- Which character was used for decimal points?
- Which character was used for missing data (between two seperators where there's no value)?

:::{.callout-note title="Preview File"}
You can preview the data file in different ways, such as:

- Opening it with a text editor;
- Clicking the file name and selecting "View File" in the RStudio *File Pane*;
- Or by using the terminal if you are using Unix-based system: `head -n2 data_anonym_struc1_noise.csv` (to view the first 2 lines) or `more data_anonym_struc1_noise.csv` (to scroll through the file and quit by typing `q`), which is **recommended for large files**.

:::

2. Import the `data_anonym_struc1_noise.csv` into RStudio, you can use either:

- the `read_csv2()` from the package {`readr`} (`?readr::read_csv2`), or
- use the click-button way and copy-paste the code in your script.

**Don't forget to use/select the appropriate parameters to make sure you import correctly the data.**

Name the data as `data1`.
Convert your imported data to tibble format if it's not the case.

What is the data dimension?

```{r}
data1 <- read_csv2(
  file = "../exos_data/data_anonym_struc1_noise.csv",
  locale = locale(decimal_mark = ",")
)

# or use the read_delim() with appropriate parameters
# data1 <- read_delim(
#   "../exos_data/data_anonym_struc1_noise.csv", 
#   delim = ";", locale = locale(decimal_mark = ",")
# )
```

3. Show the first 10 columns of your data.

```{r}
# classic way
data1[, 1:10]

# use select()
select(data1, 1:10)
```

4. Rename the first column as `age`.

```{r}
data1 <- data1 |>
  rename(age = ...1)

```

:::{.callout-tip appearance="simple"}
- What can we do with the current data format?
- If we want to calculate the average expression value of a given gene by sex or by age,
how should the data be organized?
:::


### Reshape the Data

5. Reshape the dataset to a longer format using `pivot_longer()`, so that for each mouse:

- The first column shows the age
- The second column shows the sample ID (*e.g.*, 1MA, 1MB, ...)
- The third column shows the gene expression value

What are the columns to be included to pivot into longer format?

```{r}
# use column index
data1_long <- data1 |>
  pivot_longer(cols = -1, names_to = "id", values_to = "value")

# use column name
# data1_long <- data1 |>
#   pivot_longer(cols = -age, names_to = "id", values_to = "value")

data1_long 
```

6. Add a column `struc` which contains the name of the measured structure `s1`.

```{r}
data1_long <- data1_long |>
  mutate(struc = "s1")
data1_long
```

7. Extract information about gene, sex and animal from the column `id` using the `extract()` function. Name the new columns as "gene_id", "sex" and "animal".

::: {.callout-tip collapse="true" title="Hints"}
Find the patterns for the extraction.
You can use AI to help you to write the regular expression.
:::

```{r}
# check how id (the colnames before reshape) was constructed.
head(data1_long$id)
tail(data1_long$id)

data1_long <- data1_long |>
  extract(
    col = id,
    into = c("gene_id", "sex", "animal"),
    regex = "([0-9]+)([MF])([A-F])"
  )
```

If you want to write all previous codes together:

```{r}
#| eval: false

data1_long <- read_csv2(
    file = "../exos_data/data_anonym_struc1_noise.csv",
    locale = locale(decimal_mark = ",")
  ) |>
  rename(age = ...1) |> # not necessary if the next line "cols" based on index
  pivot_longer(cols = -1, names_to = "id", values_to = "value") |>
  mutate(struc = "s1") |>
  extract(
    col = id,
    into = c("gene_id", "sex", "animal"),
    regex = "([0-9]+)([MF])([A-F])"
  )
```

Now, the data is ready for downstream analysis.

### Manipulate the Data

For question 8 to 11, let's focus on gene 1 from the data.

8. At age of 10 days, which animal has the highest expression value for gene 1 overall?

```{r}
# overall
data1_long |>
  filter(gene_id == "1" & age == 10) |>
  arrange(desc(value)) |>
  head(1)
```

9. Is there any missing value for gene 1?
If yes, how to remove lines with NA?

```{r}
table(is.na(data1_long$value[data1_long$gene_id == "1"]))

data1_long |>
  filter(gene_id == "1") |>
  drop_na(value)
```

10. After removing NAs, how many animals are there for each sex in gene 1?

```{r}
data1_long |>
  filter(gene_id == "1") |>
  drop_na(value) |>
  group_by(sex) |>
  count()
```

11. Summarize the median, mean, and standard deviation of gene 1 expression for both sexes.

```{r}
data1_long |>
  filter(gene_id == "1") |>
  drop_na(value) |>
  group_by(sex) |>
  summarise(
    median_gene1 = median(value),
    mean_gene1 = mean(value),
    sd_gene1 = sd(value)
  )
```

### Explore the Data

What kind of analysis would you like to perform with this data?

In statistics, it's common to begin by exploring the dataset as a whole and visualizing the relationships between different variables.
The function `ggpairs()` from the {[`GGally`](https://ggobi.github.io/ggally/index.html)} package
is useful for creating a matrix of scatter plots to examine the relationships between each pair of variables.
First, let's install the {`GGally`} package if it is not already installed.

```{r}
#| eval: false

install.packages("GGally")
```

12. Load the package and take a look at the help page for `ggpairs()`.
How will you reshape the `data1_long` to provide the necessary data for the `[ggpairs()](https://ggobi.github.io/ggally/articles/ggpairs.html)` function?

```{r}
#| eval: false

library(GGally)
?ggpairs
```

```{r}
data1_wider <- data1_long |>
  pivot_wider(
    names_from = "gene_id",
    values_from = "value",
    names_prefix = "gene" # to avoid name starts with number
  )

data1_wider
```

To save space, we will focus on examining the relationship between age and the first 5 genes.

13. What did you observe from these scatter plots?

```{r}
#| echo: false

library(GGally)
```

```{r}
#| fig-width: 12
#| fig-height: 12
#| warning: false

ggpairs(select(data1_wider, age, gene1:gene5))
```


The expression between gene 1 and 2 is similar, with a high coefficient of correlation.
We observe the same trend between the expression of genes 3 and 4.

In addition, it seems that there are two groups of mice that express genes 3 and 5 in a similar way.

14. Calculate the correlation between gene 1 and 2. (`?cor`)

```{r}
cor(
  x = data1_wider$gene1,
  y = data1_wider$gene2,
  use = "na.or.complet"
) # by default use the pearson's method
```


15. Draw a scatter plot using {`ggplot2`} to show the expression levels of genes 3 and 5.
Color the points by different categorical variables that we have, *i.e.*, age, sex, and animal.

Is there any categorical variable that can explain the groups we observed in the figure?

```{r}
p_age <- ggplot(data = data1_wider, aes(x = gene3, y = gene5)) +
  geom_point(aes(color = age)) +
  labs(title = "Expression Level of Genes 3 and 5") +
  theme_light()
p_age

p_sex <- ggplot(data = data1_wider, aes(x = gene3, y = gene5)) +
  geom_point(aes(color = sex)) +
  labs(title = "Expression Level of Genes 3 and 5") +
  theme_light()
p_sex

p_animal <- ggplot(data = data1_wider, aes(x = gene3, y = gene5)) +
  geom_point(aes(color = animal)) +
  labs(title = "Expression Level of Genes 3 and 5") +
  theme_light()
p_animal
```


---

#### Good job! ðŸ‘ðŸ‘ You've made great progress in mastering data manipulation techniques.


