---
title: "Week 6 - Homework"
format: 
  html: default
date: "2026-01-07"
date-format: " "
categories: 
  - week06
  - homework
editor_options: 
  chunk_output_type: console
---

For this homework, we will use the [Palmer Penguins dataset](https://allisonhorst.github.io/palmerpenguins/).
It contains information about three penguin species, living in the Palmer Archipelago.
For each penguin, measurements include:

- Bill length and bill depth (in mm)
- Flipper length (in mm)
- Body mass (in grams)
- Sex and the island where it was observed
- Study year

The goal is to explore, manipulate and visualize these data in R while practicing the concepts learned in class.


0. Set up your workspace.

Create a new R Project called `penguins_analysis` and save the dataset file [penguins.csv](https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv) inside a folder named `data` in your project.


1. Import the `penguins.csv` dataset, name the imported data as `penguins`.

```{r}

```

2. Insepct the imported data: 

- How many rows and columns are there?
- Recode the `species`, `island` and `sex` as factors

```{r}

```

3. Which species contains the most data?

```{r}

```

4. Extract penguins of "Gentoo" species which live on the "Biscoe" island.

```{r}

```

5. Create a new column `bm_kg` which stores the body mass in kilogram.

```{r}

```

6. Which species shows the highest average body mass in kilogram?

```{r}

```

7. Use scatter plot to visualise the body mass (in kg) and the bill length:

- Color the point by species.
- Increase the point size to 2.5.
- Replace the axis' label by "Body Mass (kg)" and "Bill Length (mm)".
- Remove the legend title.
- Use `theme_minimal()`.
- Put the legend on the top of the figure.

```{r}

```

What is your observation for Adelie and Chinstrap penguins?


8. Based on your observation from question 7, test if your hypothesis is statistically valid (using significance level at 5%). 

::: {.callout-tip collapse="true" title="Hints"}
- Check the distribution of body mass or bill length,
test the normality of data distribution of each species.
- If both group are normally distributed, use a t-test (`?t.test`);
If one or both groups are not normally distributed, use a Wilcoxon rank-sum test.
- Report the p-value and state your conclusion in one sentence.
:::

```{r}

```

---

The homework correction is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week06/.r_w06_work_corr.R)