---
title: "Week 6 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2026-01-06"
date-format: " "
categories: 
  - week06
  - exercise
editor_options: 
  chunk_output_type: console
execute: 
  error: true
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week06/r_w06_exos.R)

## Goals

- Hands-on challenge: “Fix the Code”
- Mini data analysis project
  - Data cleaning
  - Simple analysis
  - Visualisation

## "Fix the Code" Challenge

1. Fix the data frame error.

```{r}
data <- data.frame(
  gene = c("BRCA1", "TP53", "MYC"),
  expression = c(10.5, 8.2, 12.7)
  condition = c("Tumor", "Normal", "Tumor")
)
```

```{r}
#| code-fold: true
#| code-summary: "Correction"

data <- data.frame(
  gene = c("BRCA1", "TP53", "MYC"),
  expression = c(10.5, 8.2, 12.7),
  condition = c("Tumor", "Normal", "Tumor")
)
```

2. Fix error when using the `mean()` function.

```{r}
counts <- data.frame(
  sample1 = c(1, 2, 3),
  sample2 = c(1, 1, 3),
  sample3 = c(0, 1, 2),
  row.names = paste0("gene", 1:3)
)
counts

mean(counts["gene3", ])
```

```{r}
#| code-fold: true
#| code-summary: "Correction"

mean(unlist(counts["gene3", ]))
# or
mean(as.matrix(counts)["gene3", ])
```

3. Fix error in condition.

```{r}
x <- 10
if (x = 5) {
  print("x is 5")
}
```

```{r}
#| code-fold: true
#| code-summary: "Correction"

x <- 10
if (x == 5) {
  print("x is 5")
}
```


4. Fix error in ggplot2.
The goal is to show petal length with a boxplot for each species.

```r
ggplot(iris, aes(x = Sepal.Length, y = Species))
geom_boxplot()
```
```{r}
#| echo: false
ggplot(iris, aes(x = Sepal.Length, y = Species))
geom_boxplot()
```

```{r}
#| code-fold: true
#| code-summary: "Correction"

library(ggplot2) # load the package before use
ggplot(iris, aes(x = Sepal.Length, y = Species)) + # need "+" between layers
  geom_boxplot()
```

5. Fix error in ggplot2. The aim is to draw boxplot for each group.

```{r}
# simulate data for two groups of samples.
set.seed(1)
data <- data.frame(
  gp = rep(1:2, each = 20),
  value = c(rnorm(20), rnorm(20, mean = 5))
)
str(data)
# draw boxplot by group.
ggplot(data, aes(x = gp, y = value)) + 
  geom_boxplot()
```

```{r}
#| code-fold: true
#| code-summary: "Correction"

ggplot(data, aes(x = factor(gp), y = value)) + 
  geom_boxplot()
```

6. Fix code in data filtration.

```{r}
# try to keep rows where the value is smaller than -0.5
data[data$value<-0.5, ]
```

```{r}
#| code-fold: true
#| code-summary: "Correction"

str(data) # the original data was changed!

# rebuild the data frame
set.seed(1)
data <- data.frame(
  gp = rep(1:2, each = 20),
  value = c(rnorm(20), rnorm(20, mean = 5))
)

data[data$value < -0.5, ]
```

## Mini Data Project

A researcher has collected some gene expression data from 12 samples.
However, some expression values are missing.
Please help the researcher to clean the data and to performs some basic analyses.

```{r}
# Simulated dataset with missing values
data <- data.frame(
  sample_id = paste0("sample", 1:12),
  expression = c(
    10.2, 15.2, NA, NA, 9.4, 18.1,
    8.9, 16.0, 10.5, 15.5, 11.5, 13.4
  ),
  sample_group = rep(c("Control", "Case"), times = 6)
)

# Show the dataset
data
```

Tasks:

1. Find missing values.
Which rows contain missing values?
→ Hint: Use `is.na()`

```{r}
#| code-fold: true
#| code-summary: "Correction"

# Check which values are missing
is.na(data$expression)

# Show only rows with missing values
data[is.na(data$expression), ]

## or find the index of NAs
which(is.na(data$expression))
data[which(is.na(data$expression)), ]
```

2. Remove rows with missing values.
Create a new dataset without missing values.

```{r}
#| code-fold: true
#| code-summary: "Correction"

# Remove rows where expression is NA
data_clean <- data[!is.na(data$expression), ]

# Show the cleaned dataset
data_clean
```

3. Basic summary statistics

- What is the mean expression level (after removing missing values)?
- What is the max and min expression?

```{r}
#| code-fold: true
#| code-summary: "Correction"

# Compute summary statistics
mean_expression <- mean(data_clean$expression)
mean_expression

max_expression <- max(data_clean$expression)
max_expression

min_expression <- min(data_clean$expression)
min_expression

## or use the range() function for the min and max
range(data_clean$expression)
```

4. Find the average expression (`mean()`) and the standard deviation (`sd()`) 
for each sample group (Control *vs.* Case)

```{r}
#| code-fold: true
#| code-summary: "Correction"

# Compute mean and sd of expression for each sample group
expr_ctrl <- data_clean$expression[data_clean$sample_group == "Control"]
mean(expr_ctrl)
sd(expr_ctrl)

expr_case <- data_clean$expression[data_clean$sample_group == "Case"]
mean(expr_case)
sd(expr_case)
```

5. Use `data_clean` to draw a graph you have already seen, *e.g.*: box plots, scatter plots, *etc.*

```{r}
#| code-fold: true
#| code-summary: "Correction"

## a boxplot to show if "Case" has higher/lower expression than "Control."
ggplot(
  data_clean,
  aes(x = sample_group, y = expression, fill = sample_group)
) +
  geom_boxplot() +
  labs(
    title = "Expression Levels by Group",
    x = NULL,
    y = "Expression"
  ) +
  theme(legend.position = "none")

## Violin plot to show variability and density of expression levels
ggplot(
  data_clean,
  aes(x = sample_group, y = expression, fill = sample_group)
) +
  geom_violin() +
  labs(
    title = "Violin Plot - Expression Distribution per Group",
    x = NULL,
    y = "Expression"
  ) +
  theme(legend.position = "none")

## Spotting individual variations
ggplot(
  data_clean,
  aes(x = sample_id, y = expression, color = sample_group)
) +
  geom_point(size = 3) +
  labs(
    title = "Scatter Plot - Expression per Sample",
    x = NULL,
    y = "Expression",
    color = NULL
  )

## better solution
ggplot(
  data_clean,
  aes(
    # recode variable as factor with the wanted order
    x = factor(sample_id, levels = sample_id),
    y = expression,
    color = sample_group
  )
) +
  geom_point(size = 3) +
  labs(
    title = "Scatter Plot - Expression per Sample",
    subtitle = "Improved Version :)",
    x = NULL,
    y = "Expression",
    color = NULL
  ) +
  theme(
    # rotate x-axis' text to avoid overlap
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```


6. **To go futhur**: Let's visualise the average expression of each group using barplot,
with the help of ChatGPT (or any other AI tool).

6a. Prepare data for barplot with error bars.
We need to reorganize the data in a dataframe with 2 rows and 3 columns:

- the column `group` contain the name of each group 
- the column `mean` contain the mean expression in  of each group 
- the column `mean` contain the standard deviations


```{r}
#| code-fold: true
#| code-summary: "Correction"

gg_data <- data.frame(
  group = c("Control", "Case"),
  mean_val = c(mean(expr_ctrl), mean(expr_case)),
  sd_val = c(sd(expr_ctrl), sd(expr_case))
)

gg_data
```

6b. Draw the bar plot:

- Plot bars for mean expression (`geom_bar(stat = "identity")`)
- Add error bars for standard deviation (`geom_errorbar()`)
- Change the aesthetic apsects as you want, *e.g.*: color, title, legend, *etc.*

```{r}
#| code-fold: true
#| code-summary: "Correction"

p_bar <- ggplot(data = gg_data, aes(x = group, y = mean_val)) +
  geom_bar(stat = "identity", fill = "orange", width = 0.5) # bar plot
p_bar

p_bar_error <- p_bar +
  geom_errorbar( # add errorbar
    aes(
      ymin = mean_val - sd_val, 
      ymax = mean_val + sd_val
    ), 
    width = 0.2 # smaller width of error bar
  )

# change labels and theme
p_bar_error +
  labs(
    title = "Mean Gene Expression by Sample Group",
    subtitle = "The error bars represent the standard deviation.",
    x = NULL, y = "Mean Expression Level"
  ) +
  theme_minimal()
```


6c: What if we want to add the expression level of each sample to the bar plot?

Hint: Add another layer for drawing points (`geom_point`), using the data frame that contains the individual data (`data_clean`).

```{r}
#| code-fold: true
#| code-summary: "Correction"

# change labels and theme
p_bar_error +
  labs(
    title = "Mean Gene Expression by Sample Group",
    subtitle = "The error bars represent the standard deviation.",
    x = NULL, y = "Mean Expression Level"
  ) +
  theme_minimal() +
  geom_point( # add the layer of points
    data = data_clean, # use the individual data
    aes(x = sample_group, y = expression), # map axes
    position = position_jitter( # add random noise to avoid overlap between points
      width = 0.1, height = 0, # allow the points to be spread across a small horizontal range (width = 0.1),
      # while keeping the value on y-axis fixed (height = 0)
      seed = 1
    ) 
  )
```



---

#### Good job! 👏👏 You've taken your first big steps into R, and you're off to a great start, keep it up! 


