---
title: "Week 8 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2026-01-27"
date-format: " "
categories: 
  - week08
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/week08/r_w08_exos.R)

## Goals

## Today's Goals

- Create more complex figures
- Learn to write custom functions to automate repetitive visualization tasks


## Import Data

We will re-use the gene-level differential expression (DE) analysis results ([`toy_DEanalysis.csv`](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/exos_data/toy_DEanalysis.csv)) seen on session 4.
It was obtained by comparing SET1 samples to WT samples using count data from [read-counts.csv](https://raw.githubusercontent.com/InforBio/ioc-r-2026/refs/heads/main/ioc_r/exos_data/read-counts.csv).

- Import the DE results and name it as `de_res`.
- Import the count data and name it as `counts`.

```{r}
library(tidyverse)

de_res <- read_csv("../exos_data/toy_DEanalysis.csv")
counts <- read_csv("../exos_data/read-counts.csv")
```


1. Set `theme_minimal()` as the default theme, and a color palette for WT ("#336872") and SET1 ("#EF7B30") samples.

```{r}
theme_set(theme_minimal())
colors <- c("WT" = "#336872", "SET1" = "#EF7B30")
```

2. Pick the up-regulated gene with the smallest adjusted p-value.
Make a boxplot of the expression counts across WT and SET1 samples,
using:
- `geom_boxplot()` and `geom_points()` 
- custom colors by group

Store the final plot as `p`.

```{r}
# find the top up-regulated gene
filter(de_res, log2FoldChange > 0) |>
  arrange(padj)

the_gene <- "LOH1"

# extract expression data and reshape for boxplot
df_plot <- filter(counts, Feature == the_gene) |>
  select(contains(names(colors)), -contains("RRP6")) |>
  pivot_longer(everything()) |>
  mutate(group = sub("(.*)\\.([0-9]+)", "\\1", name))
df_plot

# draw the plot
p <- ggplot(df_plot, aes(x = group, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    aes(color = group),
    position = position_jitter(width = 0.3, seed = 8)
  ) +
  scale_color_manual(values = colors) +
  labs(x = NULL, title = paste(the_gene, "gene")) +
  theme(legend.position = "none")
p
```

3. Label the top 3 expressed samples in each group in the previous plot,
using `geom_repel_label()` of {ggrepel} package.

Hint:

- Use `group_by()` to manipulate the data by group.
- Use `rank()` to get the rank of values. (`?rank`)

Store the final plot as `p_label`.

```{r}
# E.g.: rank sample based on values by group
df_plot |>
  group_by(group) |>
  mutate(rank = rank(value))


# label the top 3 expressed sample in each group
df_plot <- df_plot |>
  group_by(group) |>
  mutate(label = ifelse(rank(desc(value)) <= 3, name, NA))


# redo the figure with the updated data frame and add labels
p_label <- ggplot(df_plot, aes(x = group, y = value)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    aes(color = group),
    position = position_jitter(width = 0.3, seed = 8)
  ) +
  scale_color_manual(values = colors) +
  labs(x = NULL, title = paste(the_gene, "gene")) +
  theme(legend.position = "none") +
  ggrepel::geom_label_repel(
    aes(label = label),
    position = position_jitter(width = 0.3, seed = 8),
    min.segment.length = 0
  )
p_label
```


4. Based on previous figure, add the adjusted p-value by using `stat_pvalue_manual()` from the {ggpubr} package. (`?stat_pvalue_manual`)

Store the final plot as `p_signif`.

```{r}
library(ggpubr)

pvalue_df <- data.frame(
  group1 = "WT",
  group2 = "SET1",
  p.adj = signif(de_res$padj[de_res$gene_name == the_gene], digits = 3), # The p-value you want to display
  y.position = 230 # The y-coordinate where the annotation will be placed
)

p_signif <- p_label + stat_pvalue_manual(
  data = pvalue_df,
  y.position = 250
)
p_signif

```


5. Create a horizontal bar plot showing the log2 fold change for each gene and
highlight those with an adjusted p-value (`padj`) of less than 0.05.

Hints: 

- Create a new column to store the rank of each gene based on its log2FoldChange value.
- Create a new column to indicate if the gene satistifies the criteria or not.

Store the final plot as `p_lfc`.

```{r}
#| fig-height: 8

de_res <- de_res |>
  arrange(log2FoldChange) |>
  mutate(
    significant = ifelse(padj < 0.05, "yes", "no"),
    gene_name_fct = factor(gene_name, levels = gene_name)
  )

p_lfc <- ggplot(
  de_res,
  aes(x = log2FoldChange, y = gene_name_fct, fill = significant)
) +
  geom_col() +
  scale_fill_manual(values = c("no" = "grey80", "yes" = "red")) +
  labs(x = "log2 Fold Change", y = NULL, fill = "p.adjust < 0.05") +
  theme(legend.position = "inside", legend.position.inside = c(0.2, 0.9))
p_lfc
```

6. Create a function to plot boxplots as `p` (of question 2) for gene of choice.
Test your function for the gene *PIR3*.
Store the figure as `p2`.

```{r}
draw_boxplot <- function(
  df_counts = counts,
  the_gene,
  colors = c("WT" = "#336872", "SET1" = "#EF7B30")
) {
  df_plot <- filter(df_counts, Feature == the_gene) |>
    select(contains(names(colors)), -contains("RRP6")) |>
    pivot_longer(everything()) |>
    mutate(group = sub("(.*)\\.([0-9]+)", "\\1", name))

  p <- ggplot(df_plot, aes(x = group, y = value)) +
    geom_boxplot(outlier.shape = NA) +
    geom_point(
      aes(color = group),
      position = position_jitter(width = 0.3, seed = 8)
    ) +
    scale_color_manual(values = colors) +
    labs(x = NULL, title = paste(the_gene, "gene")) +
    theme(legend.position = "none")

    return(p)
}

p2 <- draw_boxplot(df_counts = counts, the_gene = "FAR1")
p2
```

7. Create a patchwork layout for `p_lfc`, `p_signif`, and `p2`.
Place `p_lfc` on the left, and arrange the two boxplots (`p_signif` and `p2`) on the right using the `|` or `/` operators.

```{r}
#| fig-height: 8
#| #| fig-width: 10

library(patchwork)

p_lfc | p_signif / p2 

```

## Bonus Question

8. Explore the `wrap_plots()` function from {`patchwork`}. Try to:

- Recreate the same figure arrangement as in the previous question, this time using the "design" argumen.
- Add panel labels ("A", "B", "C") to the plots with the `plot_annotation()` function.

```{r}
#| fig-height: 8
#| fig-width: 10

wrap_plots(p_lfc, p_signif, p2, design = "AB\nAC") +
  plot_annotation(tag_level = "A")
```

---

#### Good job! ðŸ‘ðŸ‘ You've made great progress in mastering data manipulation techniques.


