---
title: "Week 4 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2025-01-28"
date-format: " "
categories: 
  - week04
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/week04/r_w04_exos.R)

## Goals

- Understand and use operators to filter data with precision.
- Understand function's basic structure, learn to use new functions, write functions to automate tasks.

## Import Data

A gene-level differential expression (DE) analysis was performed to compare SET1 samples to WT samples using data from `read-counts.csv`.

The analysis results are available via this [link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/exos_data/toy_DEanalysis.csv).

- Please donwload the result file and upload it to your data folder.
- Import the data using the `read.csv()` function.
(See the documentation with `?read.csv`) 
Name the imported results `de_res`.

```{r}
de_res <- read.csv(
  file = "../exos_data/toy_DEanalysis.csv",  # replace the path with your own
  header = TRUE
)
```


## Exercises

1. Check the structure of `de_res` using an appropriate R function. What are the dimensions?

```{r}
str(de_res)
```

The result is a data frame with `r nrow(de_res)` rows and `r ncol(de_res)` columns.

The result contains following columns:

- `gene_name`: gene name
- `baseMean`: mean of normalized counts for all samples
- `log2FoldChange`: log2 fold change
- `lfcSE`: standard error
- `stat`: Wald statistic
- `pvalue`: Wald test p-value
- `padj`: adjusted p-values (Benjamini-Hochberg procedure)

2. Filter the rows where the gene has a log2 fold change (`log2FoldChange`) greater than 0.5.

```{r}
de_res[de_res$log2FoldChange > 0.5, ]
```

3. Filter the rows where the gene has a log2 fold change smaller than -0.5.

```{r}
de_res[de_res$log2FoldChange < -0.5, ]
```

4. Filter the rows where the gene has a log2 fold change greater than 0.5 or smaller than -0.5.

```{r}
de_res[de_res$log2FoldChange > 0.5 | de_res$log2FoldChange < -0.5, ]

## Bonus: we can test the absolute value of log2FoldChange to simplify condition
abs(c(0.5, -0.5)) # how abs() works
de_res[abs(de_res$log2FoldChange) > 0.5, ]
```

5. Filter the rows where the gene has a log2 fold change greater than 0.5 and adjusted p-value (`padj`) smaller than 0.05.

```{r}
de_res[de_res$log2FoldChange > 0.5 & de_res$padj < 0.05, ]
```

::: {.callout-tip title="Stats Time!"}
**Multiple Tests Correction**

<ul>
<li>Why multiple testing is a problem?

When performing multiple statistical tests,
the probability of making at least one Type I error (false positive) increases
with the number of tests.

For instance, if we perform 100 independent tests with a significance level ($\alpha$) of 5%,
the chance of incorrectly rejecting at least one null hypothesis is no longer 5%, but much higher. This is because the errors accumulate across the tests.

If we do 100 tests simultaneously and set and use $\alpha$ at 0.05,
the probability to do at least one error is:

$$
\begin{aligned}
P(\text{at least 1 significant result by chance}) &= 1- P(\text{non significant results}) \\
&= 1 – (1 - 0.05)^{100} \\
&= 0.99
\end{aligned}
$$

</li>

<li>Multiple test correction

To address this issue and control the overall Type I error rate,
statistical corrections like the Bonferroni correction or
False Discovery Rate (FDR) adjustments are commonly used in multiple testing scenarios.

  <ul>
    <li> **Bonferroni** correction: adjust the significance threshold ($\alpha$) to account for the number of tests (Ntest) being performed, *i.e.*, $\alpha_{adjusted}= \frac{\alpha}{\text{Ntest}}$</li>
  <p>
$$
\begin{aligned}
P(\text{at least 1 significant result by chance}) &= 1 – (1 - \frac{0.05}{100})^{100} \\
&= 0.049
\end{aligned}
$$
  </p>
    <li> **FDR** (False discovery rate): control the proportion of false positive amongst all significant results, *e.g.*: Benjamini-Hochberg (BH) procedure.</li>
  </ul>
  </li>
</ul>

:::

6. Extract results for these genes: *RNR1*, *PIR3*, *SRP68*.

```{r}
de_res[de_res$gene_name %in% c("RNR1", "PIR3", "SRP68"), ]
```

7. Use `ifelse()` to categorize genes. 
Add a new column, `gene_category`, that assigns categories:

- "up" if `log2FoldChange > 0.5`.
- "down" if `log2FoldChange < -0.5`.
- "neutral" otherwise.

```{r}
de_res[["gene_category"]] <- ifelse(
  test = de_res[["log2FoldChange"]] > 0.5,
  yes = "up",
  no = ifelse(
    test = de_res[["log2FoldChange"]] < -0.5,
    yes = "down",
    no = "neutral"
  )
)
```

8. Use `table()` to count the occurrences of each gene category.

```{r}
table(de_res[["gene_category"]])
```


## Bonus Questions

9. Write a function to automate "de_res" filtering for genes
with a p-value less than or equal to a custom cutoff.

```{r}
filter_p <- function(cutoff = 0.05) {
  de_res[de_res$pvalue <= cutoff, ]
}
filter_p(0.01)
filter_p(0.0001)
```

10. Based on the function created in question 9,
modify the function to allow output ordered by any desired column in `de_res`.

Hints: You need an extra parameter to specify the wanted column and
another parameter to fix the cutoff.

```{r}
filter_col <- function(col = "pvalue", cutoff = 0.05) {
  de_res[de_res[[col]] <= cutoff, ]
}
filter_col(cutoff = 0.0001)
filter_col(col = "log2FoldChange", cutoff = -1)
```


::: {.callout-tip}
**Ensembl Data Base**

[Ensembl](http://www.ensembl.org/index.html)
is a comprehensive genome database that provides detailed information on genes and
their annotations across a wide range of species (humain, mouse, zebrafish, *etc.*).
It integrates genomic data with tools like BioMart,
making it easy to query and extract information such as gene names, coordinates, functions,
and orthologs for research purposes.
:::

11. A yeast gene annotation file was obtained from the Ensembl data base.
This file can be donwloaded [here](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/exos_data/yeast_gene_annot.csv).

Import the data and add the annotation to the `de_res` data frame using `merge()` function.

```{r}
annot <- read.csv(
  "../exos_data/yeast_gene_annot.csv", # replace the path by your own
  header = TRUE
)

de_res <- merge(de_res, annot, by = "gene_name", all.x = TRUE)
head(de_res)
```


---

#### Bravo! 🎉 You've learned the basics of R, and you're already making great progress, keep it up!


