---
title: "Week 4 - Homework"
format: 
  html: default
date: "2025-01-29"
date-format: " "
categories: 
  - week04
  - homework
editor_options: 
  chunk_output_type: console
---

Write your code directly in a Quarto document.

To create a Quarto document: go to **File** -> **New File** -> **Quarto Document**, then click **Create**.

1. Import the `read-counts.csv` file.

```{r}

```

*Quick reminder*:
this data file contains gene expression values of samples from four groups,
sample names are prefixed by "WT.", "SET1.", "SET1.RRP6." and "RRP6.".
Each group has 10 samples.

2. Create a function to calculate the average gene expression across samples
by selected group. Test your function on WT samples.

```{r}

```

3. Use function created in question 2 to calculate the average for
all sample groups, store the results in a list.
Show the first 5 average gene expression of SET1.RRP6 samples.

```{r}

```

4. Transform the list obtained in question 3 to a data frame using `as.data.frame()`.
Show the head lines of your data frame.

```{r}

```

:::{.callout-tip}
A data frame can be considered as a list of equal-length vectors.
:::

5. What are the genes having an average greater than 10000 in WT and SET1 samples?
Compare if there are genes in common using learned operator or the `intersect()` function.

```{r}

```

6. Create a function to check if the average expression of each sample group
is normally distributed (`?shapiro.test()`) using significance level at 5%.
If `TRUE`, draw directly a histogram (`?hist()`) for the values.
Otherwise, draw a histogram for the log-transformed values.
And return the p-value of normality test at the end.

Test your function on average expression of RRP6 samples.

```{r}

```


7. Click "Render" to generate your Quarto report.

The homework correction is available here:
[link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/week04/.r_w04_work_corr.qmd)